### 2019-02-18 refinement-parsers.tex
Initial commit: outline of parser article

\documentclass{article}

\title{Verified parsers using the refinement calculus and algebraic effects}
\author{Tim Baanen \and Wouter Swierstra}

\begin{document}
There are various ways to write a parser in functional languages, for example using parser combinations.
How do we ensure these parsers are correct?
Previous work has shown that predicate transformers are useful for verification of programs using algebraic effects.
This paper will show how predicate transformers and algebraic effects allow for formal verification of parsers.

\section{Recap: algebraic effects and predicate transformers}
Algebraic effects were introduced to allow for incorporating side effects in functional languages.
For example, nondeterministic programs:
\begin{code}
ENondet = eff CNondet RNondet
\end{code}

To give specifications of programs that incorporate effects,
we can use predicate transformers.
\begin{code}
wpNondetAll : (a \to Free ENondet b) \to (a \to b \to Set) \to a \to Set
wpNondetAll = ?
\end{code}

\section{Almost parsing regular languages}
As an example of using nondeterminism,
we can define the type of regular expressions |Regex| in Agda,
and a relation |Match| on |Regex| and |String|.

Using nondeterminism as an effect, we create a function that takes a |Regex| and |String| and gives all potential matchings.
Unfortunately, we get stuck in the case of _*, since it is not immediately obvious that this terminates.
We just fail in this case.
\begin{code}
match : (r,xs : Pair Regex String) \to Free ENondet (Match r,xs)
match = ?
\end{code}

Still, we can prove that this program works, as long as the regex does not contain |_*|.
Since the type guarantees that all output is a correct match,
we only need to ensure that there is output as long as a correct match exists.
Thus, the precondition states the absence of the Kleene star and that there is some way to match,
and the postcondition does not give any restrictions on the output.
\begin{code}
pre : (r,xs : Pair Regex String) \to Set
pre (r , xs) = Par (hasNo* r) (\exists Match (r , xs))
post : (r,xs : Pair Regex String) \to Match (r,xs) \to Set
post _ _ = \top

pf : wpSpec [[ pre , post ]] \refined wpNondetAll match
pf = ?
\end{code}

\section{Combining nondeterminism and general recursion}
In order to give a regex parser that also supports the Kleene star,
we add another effect: general recursion.

Here is the corresponding predicate transformer:

The original |Free| monad is modified to support a list of effects instead of a single one,
which corresponds to taking the coproduct of the functors that |Free| is a free monad over.

Now we need give new definitions of |wp|, |wpNondet| and |wpRec|.
\begin{code}
wp : \forall {a b} \to (a \to Free [] b) \to (a \to b \to Set) \to (a \to Set)
wpNondet : \forall {a b} \to (i : ENondet \in es) \to (a \to Free es b) \to (a \to Free (delete es i) b \to Set) \to (a \to Set) -- (or make the smart constructors take (ENondet \in es) and the predicate transformers take Free (ENondet :: es) b to Free es b)?
\end{code}

\section{Recursively parsing every regular expression}

Now we are able to handle the Kleene star:

We can also show correctness in the case of |_*|.

However, in this proof we do not show termination of the parsing, so it is just a proof of partial correctness.
To prove termination, it is easier to write a new parser that refines the previous implementation.

\section{Termination, using derivatives}
We can use the Brzozowski derivative to advance the regular expression a single character, giving the function |dmatch|.

Since |dmatch| always consumes a character before going in recursion, we can easily prove it calls itself on smaller arguments.
This means that for all input values, after substituting itself in the definition enough times, we get rid of all (general) recursion.
In other words: |dmatch| terminates.

Moreover, |dmatch| is a refinement of |match|, which means it is also correct:

\section{Different viewpoints of grammars}
In classical logic, a language is no more than a set of strings, or a predicate over strings: |String \to Set|.
Constructively, such predicates (even when decidable) are not very useful: the language $\{xs | xs is a valid proof of the Riemann Hypothesis\}$ has no defined cardinality.
To make them more amenable to reasoning, we impose structure on languages, for example by giving their definition as a regular expression.
For more complicated grammars than regular expressions, we will use the abstract representation of grammars by Kasper Brink.

Alternatively, we can use the Brzozowski derivative to ensure we can step though the symbols of the language,
so we get the coinductive trie of Andeas Abel:

Our last viewpoint of grammar is a much more computational one: the list-of-succesful-parses type.

To unify these different viewpoints, we will apply algebraic effects.

\section{Parsing as effect}
While we can follow the traditional development of parsers from nondeterministic state,
algebraic effects allow us to introduce new effects,
which saves us bookkeeping issues.
The |EParser| effect has a single command, which consumes the next token from the input string, or fails if the string is empty.

An algebraic parser combines the effects of |ENondet| and |EParser|, which we will write in one predicate transformer:

This allows us to define the language of an algebraic parser: all strings such that the postcondition ``the unmatched string is empty'' is satisfied.

It is also possible to give handlers that convert algebraic parsers to a trie, or to a list-of-successes function:

\section{From abstract grammars to algebraic parsers}
The parser for the dependent grammar is similar in approach to |match|,
but we modify the type to be recursive from (A : Nonterminal) to [[ A ]].
Compare the |generateParser| function by Kasper Brink.

Partial correctness is relatively simple to show as soon as we define the semantics of grammars.

Termination is somewhat more subtle: since we call the same nonterminal repeatedly,
we cannot get rid of all recursion after enough substitutions of the definition.
Consider the language given by E \to 'a' E | 'b'.
However, we can show that substituting a fixed number of times, and then giving up using |fail|.
If each right hand side has a nonterminal, then the parser terminates in this weaker sense.

\section{Left factorisation?}
In that proof, do we need to left factorise the language? In that case, can we say something about that operation?
\end{document}

### 2020-08-21 msfp2020/msfp2020.tex
A start to the slides for MSFP 2020

\documentclass{beamer}

\usepackage{fontspec}
\usepackage{minted}
\usepackage{xspace}

\setmainfont{TeX Gyre Heros}
\setmonofont{Noto Sans Mono}
\usetheme{Madrid}

\let \parskip=\baselineskip
\setminted{fontsize=\footnotesize}

\newcommand{\Agda}[1]{\texttt{\footnotesize #1}\xspace}
\newcommand{\Free}{\Agda{Free}}
\newcommand{\match}{\Agda{match}}
\newcommand{\Nondet}{\Agda{Nondet}}
\newcommand{\Regex}{\Agda{Regex}}
\newcommand{\Set}{\Agda{Set}}

\title[Combining predicate transformer semantics]{Combining predicate transformer semantics for effects: a case study in parsing regular languages}
\author{Anne Baanen \and Wouter Swierstra}
\date{September 1, 2020}

\begin{document}

\titlepage

\begin{frame}{The history}

\emph{Predicate transformer semantics} for imperative programs have been around for 45 years.
A statement becomes a map from postcondition (a predicate on the state space) to the weakest precondition that ensures the postcondition will be satisfied.

\pause

\emph{Parser combinators} have been around in functional programming for 35 years.
Parsing becomes an effectful (monadic) operation, nondeterministically modifying state (the unparsed part of the string).

\pause

\emph{Algebraic effects} have recently gained traction in the programming language community.

\end{frame}

\begin{frame}[fragile]{Algebraic effects}

Algebraic effects separate the syntax and semantics of effects.

\begin{itemize}
	\item The syntax describes the sequencing of the primitive operations
	\item The semantics assigns meaning to these operations
\end{itemize}

\begin{minted}{Agda}
data Free (C : Set) (R : C -> Set) : Set -> Set where
	Pure : a -> Free C R a
	Op : (c : C) -> (k : R c -> Free C R a) -> Free C R a

data CNondet where
	Fail : CNondet
	Choice : CNondet
RNondet : CNondet -> Set
RNondet Fail = ⊥
RNondet Choice = Bool

Nondet = Free CNondet RNondet
\end{minted}
\end{frame}

\begin{frame}[fragile]{Semantics for algebraic effects}
\emph{Handlers} give semantics for the \Free monad naturally as a fold:

\begin{minted}{Agda}
handleList : Nondet a -> List a
handleList (Pure x) = [x]
handleList (Op Fail k) = []
handleList (Op Choice k) = k True ++ k False
\end{minted}

The generic fold that computes a predicate of type \Set:

\begin{minted}{Agda}
[[_]] : Free C R a -> ((c : C) -> (R c -> Set))
      -> (a -> Set) -> Set
[[ Pure x ]] alg P = P x
[[ Op c k ]] alg P = alg c (λ x -> [[ k x ]] alg P)
\end{minted}
\end{frame}

\begin{frame}[fragile]{Predicate transformer semantics}

A predicate transformer for commands \Agda{C} and responses \Agda{R}
is a function from postconditions of type \Agda{R -> Set} to preconditions of type \Agda{C -> Set}.
If \Agda{R} depends on \Agda{C}, this becomes:

\begin{minted}{Agda}
pt C R = (c : C) -> (R c -> Set) -> Set
\end{minted}

The type of the algebra passed to \Agda{[[\_]]} is exactly \Agda{pt C R}.
We have assigned \emph{predicate transformer semantics} to algebraic effects.

\end{frame}

\begin{frame}[fragile]{Predicate transformer semantics for \Nondet}

For nondeterminism, there are two canonical choices of predicate transformer semantics.

\Agda{ptAll} requires that all potential results satisfy the postcondition:
\begin{minted}{Agda}
ptAll Fail k = ⊤
ptAll Choice k = k True ∧ k False
\end{minted}

\Agda{ptAny} requires that there is at least one outcome that satisfies the postcondition:
\begin{minted}{Agda}
ptAny Fail k = ⊥
ptAny Choice k = k True ∨ k False
\end{minted}

\end{frame}

\begin{frame}[fragile]{Parsing regular expressions}

Consider the following syntax of regular expressions and their parse trees:
\begin{minted}{Agda}
data Regex : Set where
    Empty : Regex
    Epsilon : Regex
    Singleton : Char → Regex
    _ | _ : Regex → Regex → Regex
    _ · _ : Regex → Regex → Regex
    _ * : Regex → Regex

Tree : Regex -> Set
Tree Empty = ⊥
Tree Epsilon = ⊤
Tree (Singleton _) = Char
Tree (l | r) = Either (Tree l) (Tree r)
Tree (l · r) = Pair (Tree l) (Tree r)
Tree (r *) = List (Tree r)
\end{minted}

Let's write a parser returning these \Agda{Tree}s.
\end{frame}

\begin{frame}[fragile]{Parsing regular expressions}
A nondeterministic \Regex matcher is straightforward to write:
\begin{minted}{Agda}
match : (r : Regex) -> String -> Nondet (Tree r)
match Empty xs = Op Fail λ()
match Epsilon Nil = Pure tt
match Epsilon (_ :: _) = Op Fail λ()
match (Singleton c) xs =
    if xs = [c] then Pure c else Op Fail λ()
match (l | r) xs = Op Choice (λ b,
    if b then Inl <$> match l xs else Inr <$> match r xs)
match (l · r) xs = do
    (ys, zs) <- allSplits xs
    (,) <$> match l ys <*> match r zs
\end{minted}

\pause
... until we reach the last case:
\begin{minted}{Agda}
match (r *) xs = match (r · (r *)) xs
    -- Error: does not terminate
\end{minted}

\end{frame}

\begin{frame}[fragile]{Parsing regular expressions}
For now, we will write:
\vspace{-\baselineskip}
\begin{minted}{Agda}
match (r *) xs = Op Fail λ()
\end{minted}
\vspace{\baselineskip}

\pause

To verify our implementation, we take a specification consisting of precondition and postcondition:
\begin{minted}{Agda}
pre : Regex -> String -> Set
pre r xs = hasNo* r

post : (r : Regex) -> String -> Tree r -> Set
post r xs t = Match r xs t
\end{minted}

And check that \Agda{match} \emph{refines} this specification.

\end{frame}

\begin{frame}[fragile]{Refinement calculus}

A predicate transformer \Agda{pt1} \emph{is refined by} \Agda{pt2} if \Agda{pt₂} satisfies more postconditions than \Agda{pt₁}:
\vspace{-\baselineskip}
\begin{minted}{Agda}
_⊑_ : (pt1 pt2 : (a -> Set) -> Set) -> Set
pt1 ⊑ pt2 = ∀ P -> pt1 P -> pt2 P
\end{minted}

The relation \Agda{S ⊑ T} expresses that the program \Agda{T} is ``better'' than \Agda{S}:
\Agda{S} can be replaced with \Agda{T} everywhere, and all postconditions will still hold.

\vspace{\baselineskip}

By assigning a predicate transformer to specifications, we can also relate specifications and programs:
\vspace{-\baselineskip}
\begin{minted}{Agda}
[[_,_]] : (pre : Set) (post : a -> Set) -> (a -> Set) -> Set
[[ pre , post ]] P = pre ∧ ∀ x, post x -> P x
\end{minted}

This is the `weakest precondition' necessary so that the desired post- condition \Agda{P} holds:
\Agda{pre} should hold and any result satisfying \Agda{post} should imply the postcondition \Agda{P}.

\end{frame}

\begin{frame}[fragile]{Verification}

With these ingredients, the correctness statement of \match becomes:
\begin{minted}{Agda}
matchSound : (r : Regex) (xs : String) ->
    [[ pre r xs , post r xs ]] ⊑ [[ match r xs ]] ptAll
\end{minted}

The proof proceeds by case distinction and is uncomplicated, until we need to reason about the monadic bind operator \Agda{\_>>=\_}.

\pause

The missing ingredient is the rule of consequence:
\begin{minted}{Agda}
consequence : ∀ pt (S : Free es a) (f : a → Free es b) →
    [[ S ]] pt (λ x → [[ f x ]] pt P) ≡ [[ mx >>= f ]] pt P
\end{minted}

\end{frame}

\begin{frame}[fragile]{Adding effects}
The problem with \match is that implementing the Kleene star requires the effect of \emph{general recursion}.

We can add more effects to the free monad by choosing the command and response types from a list of \emph{effect signatures}:
\begin{minted}{Agda}
data Free (es : List Sig) : Set -> Set where
	Pure : a -> Free es a
	Op : (i : mkSig C R ∈ es) (c : C) -> (k : R c -> Free C R a) -> Free C R a
\end{minted}

We will add two new effects along with nondeterminism: general recursion and parsing.
\end{frame}

\begin{frame}[fragile]{Adding effects}
The \Agda{Rec I O} effect represents a recursive function of type \Agda{(i : I) -> O i} calling itself.
The commands are the arguments to the function and the responses are the returned values.
\begin{minted}{Agda}
Rec : (I : Set) (O : I -> Set) -> Sig
Rec I O = mkSig I O
\end{minted}
To specify the semantics of \Agda{Rec}, we need an invariant of type \Agda{(i : I) -> O i -> Set},
specifying which values of type \Agda{O i} can be returned from a call with argument \Agda{i : I}.
\begin{minted}{Agda}
ptRec inv i P = ∀ o -> inv i o -> P o
\end{minted}
\end{frame}

\begin{frame}[fragile]{Adding effects}
The \Agda{Parser} effect represents a stateful parser with one command:
advance the input string by one character.

\begin{minted}{Agda}
Parser : Sig
Parser = mkSig ⊤ (λ _ -> Maybe Char)
\end{minted}

The semantics of \Agda{Parser} are stateful: returning the next character of the input string requires keeping track of the remaining characters.
This state can be found in the \Agda{ptParser} semantics as an extra argument to the predicates.
\begin{minted}{Agda}
ptParser : ⊤ -> (Maybe Char -> String -> Set) -> String -> Set
ptParser _ P Nil = P Nothing Nil
ptParser _ P (x :: xs) = P (Just x) xs
\end{minted}

\end{frame}

\begin{frame}[fragile]{Defining a derivative-based matcher}
\begin{minted}{Agda}
dmatch : (Forall(es)) ⦃ iP : Parser ∈ es ⦄ ⦃ iND : Nondet ∈ es ⦄  -> (RecArr Regex es tree)
dmatch r = symbol >>= maybe
    (λ x -> integralTree r <$> call (hiddenInstance(∈Head)) (d r /d x))
    (if p <- matchEpsilon r then Pure (Sigma.fst p) else (hiddenConst(fail)))
\end{minted}
\end{frame}

\begin{frame}[fragile]{Termination checking}
\begin{minted}{Agda}
terminates-in : (Forall(I O es a)) (pts : PTs es) (f : (RecArr I es O)) (S : Free (Rec I O :: es) a) → Nat → Set
terminates-in pts f (Pure x)            n         = ⊤
terminates-in pts f (Op ∈Head c k)      Zero      = ⊥
terminates-in pts f (Op ∈Head c k)      (Succ n)  = terminates-in pts f (f c >>= k) n
terminates-in pts f (Op (∈Tail i) c k)   n        =
    lookupPT pts i c (λ x → terminates-in pts f (k x) n)
\end{minted}
\end{frame}

\begin{frame}[fragile]{Refining \match with \Agda{dmatch}}
\begin{minted}{Agda}
dmatchSound : ∀ r xs -> (wpMatch (match (hiddenInstance(∈Head)) (r , xs))) ⊑ (wpMatch (dmatch' (hiddenInstance(∈Head)) (r , xs)))
\end{minted}
\end{frame}

\end{document}

### 2020-08-25 msfp2020/msfp2020.tex
Update slides

@@
 \titlepage
-
-\begin{frame}{The history}
-
-\emph{Predicate transformer semantics} for imperative programs have been around for 45 years.
-A statement becomes a map from postcondition (a predicate on the state space) to the weakest precondition that ensures the postcondition will be satisfied.
-
-\pause
-
-\emph{Parser combinators} have been around in functional programming for 35 years.
-Parsing becomes an effectful (monadic) operation, nondeterministically modifying state (the unparsed part of the string).
-
-\pause
-
-\emph{Algebraic effects} have recently gained traction in the programming language community.
-
-\end{frame}
-
 \begin{frame}[fragile]{Algebraic effects}
@@
 	Op : (c : C) -> (k : R c -> Free C R a) -> Free C R a
+\end{minted}
+\end{frame}
 
+\begin{frame}[fragile]{Example: Nondeterminism}
+
+\Agda{Nondet} has two primitive operations:
+\begin{itemize}
+    \item \Agda{Choice} chooses between two values
+    \item \Agda{Fail} goes to a failure state and stops execution
+\end{itemize}
+\begin{minted}{Agda}
 data CNondet where
+	Choice : CNondet
 	Fail : CNondet
-	Choice : CNondet
+
 RNondet : CNondet -> Set
+RNondet Choice = Bool
 RNondet Fail = ⊥
-RNondet Choice = Bool
 
@@
 \emph{Handlers} give semantics for the \Free monad naturally as a fold:
-
 \begin{minted}{Agda}
@@
 handleList (Pure x) = [x]
+handleList (Op Choice k) = k True ++ k False
 handleList (Op Fail k) = []
-handleList (Op Choice k) = k True ++ k False
 \end{minted}
 
+\pause
 The generic fold that computes a predicate of type \Set:
-
 \begin{minted}{Agda}
@@
 \begin{frame}[fragile]{Parsing regular expressions}
-
-Consider the following syntax of regular expressions and their parse trees:
+To illustrate these semantics, we wrote a parser.
+The input is a regular expression and a \Agda{String}, and the output a parse tree.
 \begin{minted}{Agda}
@@
 \end{minted}
-
-Let's write a parser returning these \Agda{Tree}s.
 \end{frame}
@@
 \begin{frame}[fragile]{Parsing regular expressions}
-A nondeterministic \Regex matcher is straightforward to write:
 \begin{minted}{Agda}
@@
     if xs = [c] then Pure c else Op Fail λ()
-match (l | r) xs = Op Choice (λ b,
+
+match (l | r)       xs       = Op Choice (λ b ->
     if b then Inl <$> match l xs else Inr <$> match r xs)
+
 match (l · r) xs = do
@@
 \end{minted}
+Looks straightforward so far...
 
@@
 ... until we reach the last case:
+\vspace{-\baselineskip}
 \begin{minted}{Agda}
@@
 
-A predicate transformer \Agda{pt1} \emph{is refined by} \Agda{pt2} if \Agda{pt₂} satisfies more postconditions than \Agda{pt₁}:
+A predicate transformer \Agda{pt1} \emph{is refined by} \Agda{pt2} if \Agda{pt2} satisfies more postconditions than \Agda{pt1}:
 \vspace{-\baselineskip}
@@
 
-The relation \Agda{S ⊑ T} expresses that the program \Agda{T} is ``better'' than \Agda{S}:
+\Agda{S ⊑ T} expresses that \Agda{T} is ``better'' than \Agda{S}:
 \Agda{S} can be replaced with \Agda{T} everywhere, and all postconditions will still hold.
@@
 
-By assigning a predicate transformer to specifications, we can also relate specifications and programs:
+\pause
+
+Predicate transformers are a semantic domain where programs and specifications can be related.
 \vspace{-\baselineskip}
@@
 \end{minted}
-
-This is the `weakest precondition' necessary so that the desired post- condition \Agda{P} holds:
-\Agda{pre} should hold and any result satisfying \Agda{post} should imply the postcondition \Agda{P}.
 
@@
 \begin{frame}[fragile]{Adding effects}
-The problem with \match is that implementing the Kleene star requires the effect of \emph{general recursion}.
+The problem with \match is that implementing the Kleene star also requires the effect of \emph{general recursion}.
 
@@
     Pure : a -> Free es a
-	Op : (i : mkSig C R ∈ es) (c : C) -> (k : R c -> Free C R a) -> Free C R a
+    Op : (i : mkSig C R ∈ es) (c : C)
+         (k : R c -> Free C R a) -> Free C R a
 \end{minted}
 
-We will add two new effects along with nondeterminism: general recursion and parsing.
+We will add two new effects: general recursion and parsing.
 \end{frame}
@@
 
-The semantics of \Agda{Parser} are stateful: returning the next character of the input string requires keeping track of the remaining characters.
-This state can be found in the \Agda{ptParser} semantics as an extra argument to the predicates.
+\Agda{Parser} has stateful semantics: to return the next character, we need to keep track of the remaining characters.
+The state is the extra \Agda{String} arguments in \Agda{ptParser}.
 \begin{minted}{Agda}
@@
 
+\begin{frame}[fragile]{Extending \match}
+
+Now we can finish the definition and prove soundness unconditionally:
+\begin{minted}{Agda}
+match (r *) = Op iRec (r · (r *))
+
+matchSound : (r : Regex) (xs : String) ->
+    [[ ⊤ , post r xs ]] ⊑ [[ match r xs ]]
+\end{minted}
+
+\pause
+\match still does not terminate if \Agda{r} matches the empty string, our result is only \emph{partial correctness}.\\
+\Agda{ptRec} computes the WLP: all recursive calls immediately return.
+\end{frame}
+
 \begin{frame}[fragile]{Defining a derivative-based matcher}
+To guarantee termination, use recursion on \Agda{xs} rather than \Agda{r}.\\
+The \emph{Brzozowski derivative} \Agda{d r /d x} matches \Agda{xs} iff \Agda{r} matches \Agda{x :: xs};\\ \Agda{integralTree r : tree (d r /d x) -> tree r} ``integrates'' parse trees.
+
 \begin{minted}{Agda}
-dmatch : (Forall(es)) ⦃ iP : Parser ∈ es ⦄ ⦃ iND : Nondet ∈ es ⦄  -> (RecArr Regex es tree)
+dmatch : (r : Regex) -> Free es (tree r)
 dmatch r = symbol >>= maybe
-    (λ x -> integralTree r <$> call (hiddenInstance(∈Head)) (d r /d x))
-    (if p <- matchEpsilon r then Pure (Sigma.fst p) else (hiddenConst(fail)))
+    (λ x -> Op iRec (d r /d x) (integralTree r))
+    (if p <- matchEpsilon r
+     then Pure (Sigma.fst p)
+     else Op iND Fail λ())
+\end{minted}
+
+\pause
+
+\begin{minted}{Agda}
+dmatchSound : ∀ r xs -> [[ match r xs ]] ⊑ [[ dmatch r xs ]]
 \end{minted}
@@
 \begin{frame}[fragile]{Termination checking}
+\Agda{ptRec} gives weakest liberal precondition semantics.
+For total correctness, we should check termination.
+
+\vspace{\baselineskip}
+\Agda{terminates-in f S n} holds iff \Agda{S} terminates after calling \Agda{f} at most \Agda{n} times.
+\vspace{-\baselineskip}
+
 \begin{minted}{Agda}
-terminates-in : (Forall(I O es a)) (pts : PTs es) (f : (RecArr I es O)) (S : Free (Rec I O :: es) a) → Nat → Set
-terminates-in pts f (Pure x)            n         = ⊤
-terminates-in pts f (Op ∈Head c k)      Zero      = ⊥
-terminates-in pts f (Op ∈Head c k)      (Succ n)  = terminates-in pts f (f c >>= k) n
-terminates-in pts f (Op (∈Tail i) c k)   n        =
-    lookupPT pts i c (λ x → terminates-in pts f (k x) n)
+terminates-in : (f : (i : I) -> Free (Rec I O :: es) (O i))
+    (S : Free (Rec I O :: es) a) → ℕ → Set
+terminates-in f (Pure x)           n        = ⊤
+terminates-in f (Op ∈Head c k)     Zero     = ⊥
+terminates-in f (Op ∈Head c k)     (Succ n) =
+    terminates-in pt f (f c >>= k) n
+terminates-in f (Op (∈Tail i) c k) n        =
+    pts i c (λ x → terminates-in f (k x) n)
 \end{minted}
@@
 
-\begin{frame}[fragile]{Refining \match with \Agda{dmatch}}
+\begin{frame}[fragile]{Total correctness}
+
+Partial correctness of \Agda{dmatch} follows from the chain of refinements:
 \begin{minted}{Agda}
-dmatchSound : ∀ r xs -> (wpMatch (match (hiddenInstance(∈Head)) (r , xs))) ⊑ (wpMatch (dmatch' (hiddenInstance(∈Head)) (r , xs)))
+[[ ⊤ , post r xs ]]
+    ⊑ [[ match r xs ]]
+    ⊑ [[ dmatch r xs ]]
+    ⊑ [[ ⊤ , post r xs ]]
 \end{minted}
+together with a proof of termination:
+\begin{minted}{Agda}
+dmatchTerminates : (r : Regex) (xs : String) ->
+    terminates-in dmatch (dmatch r xs) (length xs)
+\end{minted}
+
+\end{frame}
+
+\begin{frame}{Discussion}
+
+In our paper, we illustrate how techniques from the refinement calculus can be used in functional programming.
+A distinguishing characteristic of our approach is modularity:
+we add new effects and semantics to the system as we need them.
+
+Formally verified parsers have been developed before,
+using specialized semantics to the domain of parsing.
+The modularity of predicate transformers allow us to reason about effects uniformly.
+
+Most existing approaches to recursion in parsers deal with termination syntactically.
+Separation of syntax and semantics also cleanly separates partial and total correctness.
+
 \end{frame}

### 2020-08-28 msfp2020/msfp2020.tex
Update slides
Wouter's suggestions; better theme

@@
-\documentclass{beamer}
+\documentclass[ignorenonframetext,]{beamer}
+\setbeamertemplate{caption}[numbered]
+\setbeamertemplate{caption label separator}{: }
+\setbeamercolor{caption name}{fg=normal text.fg}
+\beamertemplatenavigationsymbolsempty
+\usepackage{lmodern}
+\usepackage{amssymb,amsmath}
+\usepackage{ifxetex,ifluatex}
+\usepackage{fixltx2e} % provides \textsubscript
+%%W: moved up to allow to be overriden
+\usetheme[]{metropolis}
+\usefonttheme{serif} % use mainfont rather than sansfont for slide text
+%End
+% \ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
+%   \usepackage[T1]{fontenc}
+%   \usepackage[utf8]{inputenc}
+% % \else % if luatex or xelatex
+%   \ifxetex
+%     \usepackage{mathspec}
+%   \else
+    \usepackage{fontspec}
+  % \fi
+  % \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
+    \setmainfont[Scale=0.8]{Open Sans}
+    \setsansfont[Scale=0.8]{Open Sans}
+    \setmonofont[Mapping=tex-ansi,Scale=0.8]{Droid Sans Mono}
+%\fi
+% use upquote if available, for straight quotes in verbatim environments
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+% use microtype if available
+\IfFileExists{microtype.sty}{%
+\usepackage{microtype}
+\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
+}{}
+\newif\ifbibliography
 
-\usepackage{fontspec}
+\usepackage{color}
+\usepackage{fancyvrb}
+\newcommand{\VerbBar}{|}
+\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
+\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
+% Add ',fontsize=\small' for more characters per line
+\newenvironment{Shaded}{}{}
+\newcommand{\AlertTok}[1]{\textcolor[rgb]{1.00,0.00,0.00}{\textbf{#1}}}
+\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
+\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.49,0.56,0.16}{#1}}
+\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{#1}}
+\newcommand{\BuiltInTok}[1]{#1}
+\newcommand{\CharTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
+\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textit{#1}}}
+\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
+\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.53,0.00,0.00}{#1}}
+\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{\textbf{#1}}}
+\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.56,0.13,0.00}{#1}}
+\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{#1}}
+\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.73,0.13,0.13}{\textit{#1}}}
+\newcommand{\ErrorTok}[1]{\textcolor[rgb]{1.00,0.00,0.00}{\textbf{#1}}}
+\newcommand{\ExtensionTok}[1]{#1}
+\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{#1}}
+\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.02,0.16,0.49}{#1}}
+\newcommand{\ImportTok}[1]{#1}
+\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
+\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{\textbf{#1}}}
+\newcommand{\NormalTok}[1]{#1}
+\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.40,0.40,0.40}{#1}}
+\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{#1}}
+\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.74,0.48,0.00}{#1}}
+\newcommand{\RegionMarkerTok}[1]{#1}
+\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
+\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.73,0.40,0.53}{#1}}
+\newcommand{\StringTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
+\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.10,0.09,0.49}{#1}}
+\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
+\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
+
+%% UU Specific colors
+\definecolor{main}{RGB}{60,67,149}
+\definecolor{text}{RGB}{30,34,75}
+\definecolor{background}{RGB}{255,255,255}
+
+\definecolor{alert}{RGB}{0,159,227}
+\definecolor{example}{RGB}{60,67,149}
+
+%\definecolor{uuRed}{RGB}{192,10,53}
+%\definecolor{uuYellow}{RGB}{255,205,0}
+%\definecolor{darkGrey}{RGB}{64,64,64}
+%\definecolor{almostBlack}{RGB}{26,26,26}
+
+%\definecolor{uuCreme}{RGB}{249,230,173}
+%\definecolor{uuOrange}{RGB}{221,149,98}
+%\definecolor{uuBordeaux}{RGB}{146,30,86}
+%\definecolor{uuBrown}{RGB}{99,63,43}
+%\definecolor{uuGreen}{RGB}{99,165,147}
+%\definecolor{uuBlue}{RGB}{102,135,195}
+%\definecolor{uuDarkBlue}{RGB}{23,29,66}
+%\definecolor{uuPurple}{RGB}{82,41,128}
+
+\setbeamercolor{normal text}{fg=text, bg=white}
+\setbeamercolor{alerted text}{fg=alert}
+\setbeamercolor{example text}{fg=text}
+\setbeamercolor{frametitle}{bg=main,fg=white}
+\setbeamercolor{palette primary}{%
+   use=normal text
+  ,fg=white
+  ,bg=main
+  }
+
+\titlegraphic{\hfill\includegraphics[height=1.5cm]{img/vu.png}\includegraphics[height=1.5cm]{img/uulogo.png}}
+
+% Prevent slide breaks in the middle of a paragraph:
+\widowpenalties 1 10000
+\raggedbottom
+
+\AtBeginPart{
+  \let\insertpartnumber\relax
+  \let\partname\relax
+  \frame{\partpage}
+}
+\AtBeginSection{
+  \ifbibliography
+  \else
+    \let\insertsectionnumber\relax
+    \let\sectionname\relax
+    \frame{\sectionpage}
+  \fi
+}
+\AtBeginSubsection{
+  \let\insertsubsectionnumber\relax
+  \let\subsectionname\relax
+  \frame{\subsectionpage}
+}
+
+\setlength{\emergencystretch}{3em}  % prevent overfull lines
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+\setcounter{secnumdepth}{0}
+
 \usepackage{minted}
@@
 
-\setmainfont{TeX Gyre Heros}
-\setmonofont{Noto Sans Mono}
-\usetheme{Madrid}
+%\setmainfont{TeX Gyre Heros}
+\setmonofont[Scale=0.8]{Noto Sans Mono}
+%\usetheme{Madrid}
 
-\let \parskip=\baselineskip
-\setminted{fontsize=\footnotesize}
+% Make blocks of code appear a bit more cohesive by putting the lines closer together.
+\linespread{0.95}
 
-\newcommand{\Agda}[1]{\texttt{\footnotesize #1}\xspace}
+\newcommand{\Agda}[1]{\texttt{#1}\xspace}
 \newcommand{\Free}{\Agda{Free}}
@@
 
-\title[Combining predicate transformer semantics]{Combining predicate transformer semantics for effects: a case study in parsing regular languages}
-\author{Anne Baanen \and Wouter Swierstra}
+\title[Combining predicate transformer semantics]{Combining predicate transformer semantics for effects}
+\subtitle{A case study in parsing regular languages}
+\author{Anne Baanen \hspace{5.1em} Wouter Swierstra}
+\institute{Vrije Universiteit Amsterdam \hspace{1em} Utrecht University}
 \date{September 1, 2020}
+
+\date{}
 
@@
 
-\titlepage
+\frame{\titlepage}
+
 \begin{frame}[fragile]{Algebraic effects}
@@
 \end{itemize}
+
+In this work, we use a free monad to model effectful programs:
 
@@
 A predicate transformer for commands \Agda{C} and responses \Agda{R}
-is a function from postconditions of type \Agda{R -> Set} to preconditions of type \Agda{C -> Set}.
+is a function from postconditions of type \Agda{R -> Set} to preconditions of type \Agda{C -> Set}.\\
 If \Agda{R} depends on \Agda{C}, this becomes:
@@
 \begin{frame}[fragile]{Parsing regular expressions}
+
+We implement \match as a case distinction.
 \begin{minted}{Agda}
@@
 match Empty         xs       = Op Fail λ()
+\end{minted}
+\pause \vspace{-\baselineskip}
+\begin{minted}{Agda}
 match Epsilon       Nil      = Pure tt
 match Epsilon       (_ :: _) = Op Fail λ()
+\end{minted}
+\pause \vspace{-\baselineskip}
+\begin{minted}{Agda}
 match (Singleton c) xs       =
     if xs = [c] then Pure c else Op Fail λ()
-
+\end{minted}
+\pause \vspace{-\baselineskip}
+\begin{minted}{Agda}
 match (l | r)       xs       = Op Choice (λ b ->
     if b then Inl <$> match l xs else Inr <$> match r xs)
-
+\end{minted}
+\pause \vspace{-\baselineskip}
+\begin{minted}{Agda}
 match (l · r) xs = do
@@
 \end{minted}
-Looks straightforward so far...
-
-\pause
-... until we reach the last case:
-\vspace{-\baselineskip}
+\pause \vspace{-\baselineskip}
 \begin{minted}{Agda}
 match (r *) xs = match (r · (r *)) xs
-    -- Error: does not terminate
 \end{minted}
+\pause
+\alert{Error: \Agda{match (r *) xs} does not terminate}
 
@@
 For now, we will write:
-\vspace{-\baselineskip}
 \begin{minted}{Agda}
@@
 \end{minted}
-\vspace{\baselineskip}
 
@@
 A predicate transformer \Agda{pt1} \emph{is refined by} \Agda{pt2} if \Agda{pt2} satisfies more postconditions than \Agda{pt1}:
-\vspace{-\baselineskip}
+
 \begin{minted}{Agda}
@@
 
-\vspace{\baselineskip}
-
 \pause
@@
 Predicate transformers are a semantic domain where programs and specifications can be related.
-\vspace{-\baselineskip}
 \begin{minted}{Agda}
@@
 \begin{frame}[fragile]{Adding effects}
-The \Agda{Rec I O} effect represents a recursive function of type \Agda{(i : I) -> O i} calling itself.
+Inspired by McBride's \textit{Turing-Completeness Totally Free}, we use the \Agda{Rec I O} effect to represent a recursive function of type \Agda{(i : I) -> O i} calling itself.
 The commands are the arguments to the function and the responses are the returned values.
@@
 \begin{minted}{Agda}
-ptParser : ⊤ -> (Maybe Char -> String -> Set) -> String -> Set
-ptParser _ P Nil = P Nothing Nil
-ptParser _ P (x :: xs) = P (Just x) xs
+ptParser : (Maybe Char -> String -> Set) -> String -> Set
+ptParser P Nil = P Nothing Nil
+ptParser P (x :: xs) = P (Just x) xs
 \end{minted}
@@
 
-\vspace{\baselineskip}
 \Agda{terminates-in f S n} holds iff \Agda{S} terminates after calling \Agda{f} at most \Agda{n} times.
-\vspace{-\baselineskip}
 
@@
 In our paper, we illustrate how techniques from the refinement calculus can be used in functional programming.
+They provide a natural and uniform way to reason about effects in the setting of the \Agda{Free} monad.
+
 A distinguishing characteristic of our approach is modularity:

### 2020-08-29 msfp2020/msfp2020.tex
tweaks

@@
 \end{minted}
+
+\pause
+Let us look at \Agda{alg : (c : C) -> (R c -> Set) -> Set}'s type.
 \end{frame}
@@
 \begin{minted}{Agda}
-match (r *) xs = match (r · (r *)) xs
+match (r *) xs = match (Epsilon | r · (r *)) xs
 \end{minted}
@@
 \begin{minted}{Agda}
-consequence : ∀ pt (S : Free es a) (f : a → Free es b) →
-    [[ S ]] pt (λ x → [[ f x ]] pt P) ≡ [[ mx >>= f ]] pt P
+consequence : ∀ pt (S : Free es a) (f : a -> Free es b) ->
+    [[ S >>= f ]] pt P ≡ [[ S ]] pt (λ x -> [[ f x ]] pt P)
 \end{minted}
@@
 \begin{minted}{Agda}
-match (r *) = Op iRec (r · (r *))
+match (r *) = Op iRec (Epsilon | r · (r *)) eitherToList
 
@@
 To guarantee termination, use recursion on \Agda{xs} rather than \Agda{r}.\\
-The \emph{Brzozowski derivative} \Agda{d r /d x} matches \Agda{xs} iff \Agda{r} matches \Agda{x :: xs};\\ \Agda{integralTree r : tree (d r /d x) -> tree r} ``integrates'' parse trees.
+The \emph{Brzozowski derivative} \Agda{d r /d x} matches \Agda{xs} iff \Agda{r} matches \Agda{x :: xs}.
 
 \begin{minted}{Agda}
-dmatch : (r : Regex) -> Free es (tree r)
-dmatch r = symbol >>= maybe
-    (λ x -> Op iRec (d r /d x) (integralTree r))
-    (if p <- matchEpsilon r
+dmatch : (r : Regex) -> Free es (Tree r)
+dmatch r = Op iParse
+    λ { (Just x) -> Op iRec (d r /d x) (integralTree r)
+        Nothing  -> if p <- matchEpsilon r
                     then Pure (Sigma.fst p)
-     else Op iND Fail λ())
+                    else Op iND Fail λ() }
 \end{minted}
+
+\Agda{integralTree r : tree (d r /d x) -> tree r} ``integrates'' parse trees.
 
@@
 terminates-in f (Op (∈Tail i) c k) n        =
-    pts i c (λ x → terminates-in f (k x) n)
+    pts i c (λ x -> terminates-in f (k x) n)
 \end{minted}

### 2020-10-08 msfp2020/msfp2020.tex
Final version

@@
 
-In this work, we use a free monad to model effectful programs:
+In this work, we use a free monad to model effectful programs in Agda:
 
@@
 \begin{minted}{Agda}
-[[_]] : Free C R a -> ((c : C) -> (R c -> Set))
+[[_]] : Free C R a -> ((c : C) -> (R c -> Set) -> Set)
       -> (a -> Set) -> Set
@@
 \end{minted}
-
-\pause
-Let us look at \Agda{alg : (c : C) -> (R c -> Set) -> Set}'s type.
 \end{frame}
@@
 \begin{minted}{Agda}
-match (r *) = Op iRec (Epsilon | r · (r *)) eitherToList
+match (r *) = Op iRec (Epsilon | r · (r *))
 
